---
title: "GrandEnrichWrap"
author: "Nadav"
date: "2024-12-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Initialize

```{r}
#lapply(paste('package:',names(sessionInfo()$otherPkgs),sep=""),detach,character.only=TRUE,unload=TRUE)


library(dplyr)
library(ggplot2)

```


## Define relevant functions
```{r}

generate_run_id <- function() {
  return(format(Sys.time(), "run_%Y%m%d_%H%M%S"))
}

create_run_directory <- function(base_dir) {
  run_id <- generate_run_id()
  run_dir <- file.path(base_dir, run_id)
  dir.create(run_dir, recursive = TRUE, showWarnings = FALSE)
  return(list(run_id = run_id, run_dir = run_dir))
}

log_run <- function(master_log,parameters,data_dir,filterinng_description,
                    background_description,group_description,gene_group_count,
                    background_group_count ) {
  
  if (!file.exists(master_log)) {
    write.table(
      data.frame(
        RunID = character(),
        DateTime = character(),
        
        data_dir = character(),
        filterinng_description = character(),
        background_description = character(),
        group_description = character(),
        
        gene_group_count = numeric(),
        background_group_count = numeric(),
        
        GrandEnrich_dbs = character(),
        comp = character(),
        locORser = character(),
        minGenes = numeric(),
        uniq = character(),
        outdir = character(),
        
        #HPO specific
        TYPorALL = character(),
        HPO_dbVer = character(),
        systems = character(),
        bodyparts = character(),
        enrORdep = character(),
        directional =character(),
        
        #ORGANizer specific
        curationLevel = character(),
        FreqOfPheno =character(),
        onlySkel = character()),

        file = master_log, sep = ",", row.names = FALSE, col.names = TRUE
    )
  }
  
  log_entry <- data.frame(
    RunID = parameters$run_id,
    DateTime = format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    
    data_dir = data_dir,
    filterinng_description = filterinng_description,
    background_description = background_description,
    group_description = group_description,
    
    gene_group_count = gene_group_count,
    background_group_count = background_group_count, 
    
    GrandEnrich_dbs = paste(parameters$GrandEnrich_dbs,collapse=','),
    comp = parameters$comp,
    locORser = parameters$locORser,
    minGenes = parameters$minGenes,
    uniq = parameters$uniq,
    outdir = parameters$outdir,
    
    #HPO specific
    TYPorALL = parameters$TYPorALL,
    HPO_dbVer = parameters$HPO_dbVer,
    systems = parameters$systems,
    bodyparts = parameters$bodyparts,
    enrORdep = parameters$enrORdep,
    directional =parameters$directional,
    
    #ORGANizer specific
    curationLevel = parameters$curationLevel,
    FreqOfPheno =parameters$FreqOfPheno,
    onlySkel = parameters$onlySkel
  )
  
  write.table(log_entry, file = master_log, sep = ",", row.names = FALSE, col.names = FALSE, append = TRUE)
}

```
 

## Define parameters

```{r}
parameters = list()
base_dir = "/home/labs/davidgo/Collaboration/humanMPRA/enrichment/final_analysis"
run_dir_output = create_run_directory(paste(base_dir,'/runs',sep = ''))

parameters$GrandEnrich_dbs = c('GO_BP','KEGG','GAD_Disease')
parameters$run_id = run_dir_output$run_id
parameters$outdir = run_dir_output$run_dir

parameters$comp = "Standard with HPO filtering of skeletal"
parameters$locORser = "ser"
parameters$minGenes = 10
parameters$uniq = T

#HPO specific
parameters$TYPorALL = "ALL"
parameters$HPO_dbVer = "HPO 2024-04-26 Release"
parameters$systems = "skeleton"
parameters$bodyparts = "all"
parameters$enrORdep = "enr"
parameters$directional = F

#ORGANizer specific
parameters$curationLevel = "confident"
parameters$FreqOfPheno = "all"
parameters$DBversion = parameters$HPO_dbVer
parameters$onlySkel = F

```


## Load libraries

```{r message=FALSE, warning=FALSE}
library('dplyr')
library('tidyr')
library('ggplot2')
library('stringr')
```

## Source GrandEnrich
```{r}
grandEnrichDir <- "/home/labs/davidgo/Collaboration/backup/Lab_tools/Gokhman_Enrich/final_scripts"

source(paste(grandEnrichDir,'/TheGrandEnrich_enrich.R',sep=''))
source(paste(grandEnrichDir,'/ORGANizer_ORGANize.R',sep=''))
source(paste(grandEnrichDir,'/HPO_enrich.R',sep=''))
source(paste(grandEnrichDir,'/ORGANizer_compare.R',sep=''))
source(paste(grandEnrichDir,'/TheGrandEnrich_compare.R',sep=''))
source(paste(grandEnrichDir,'/HPO_compare.R',sep=''))


```

## Load raw data for enrichment analysis
```{r}
if(!exists('data')){
  data_dir = "/home/labs/davidgo/Collaboration/humanMPRA/top_candidates/chondrocytes/humanMPRA_annotations_v3.csv"
  data <- read.csv(data_dir)
}


```


# Data processing

## Filter the data according to preferences. Add description of filtering, which will be saved later

```{r}

filter_genes <- function(bool_str, string_str) {
  bool_vec <- str_split(bool_str, ";")[[1]] == "T"
  gene_list <- str_split(string_str, ";")[[1]]
  filtered_genes <- gene_list[bool_vec]
  paste(filtered_genes, collapse = ";")
}

filter_elite_associations <- function(num_str, string_str,within_promoter,num_associations) {
  num_str = str_split(num_str, ";")[[1]]
  within_promoter = str_split(within_promoter, ";")[[1]]
  bool_vec <-(num_str >= num_associations) | (within_promoter == 'T')
  gene_list <- str_split(string_str, ";")[[1]]
  filtered_genes <- gene_list[bool_vec]
  paste(filtered_genes, collapse = ";")
}

filterinng_description = 'filtering for elite gene - oligo association (at least 2) and for oligos with only one associated gene'



filtered_data = data%>%
  filter(DNA_counts_raw_ancestral>50)%>%
  filter(DNA_counts_raw_derived>50) %>%
  dplyr::rowwise() %>%
  mutate(Gene_symbol = filter_elite_associations(num_sources_linking_gene,Gene_symbol, within_promoter ,2))

#mutate(closest_gene = filter_elite_associations(num_sources_linking_gene, closest_gene,2))%>%
  #mutate(within_promoter = filter_elite_associations(num_sources_linking_gene, within_promoter,2))%>%
  #mutate(eRNA = filter_elite_associations(num_sources_linking_gene, eRNA,2))%>%
  #mutate(eQTL = filter_elite_associations(num_sources_linking_gene, eQTL,2))%>%
  #mutate(HiC = filter_elite_associations(num_sources_linking_gene, HiC,2))

  

  #dplyr::filter(ATACseq_peaks_fetal_chondrocytes<1)%>%
  #dplyr::filter(H3K27Ac_peaks_fetal_chondrocytes<1)


```

## Define background

```{r}

background_description = 'active'

active <- filtered_data %>%
  filter(!is.na(differential_activity))%>%
  pull(Gene_symbol)

combined_active <- paste(active, collapse = ";")
split_active <- strsplit(combined_active, split = ";")[[1]]
clean_active <- split_active[split_active != ""]
unique_active <- unique(clean_active)

background = unique_active


```

## Define group for enrichment

```{r}

group_description = 'diff active'


diff_active = filtered_data%>%
  filter(differential_activity=='True')%>%
  filter(abs(logFC_derived_vs_ancestral)>0.25)%>%
  pull(Gene_symbol)

#diff_active <- filtered_data %>%
#  filter(differential_activity == "True")%>%
#  filter(abs(logFC_derived_vs_ancestral) > log2(1.33))%>% # & abs(logFC_derived_vs_ancestral) < log2(2)) %>%
#  pull(Gene_symbol)

combined_diff_active <- paste(diff_active, collapse = ";")
split_diff_active <- strsplit(combined_diff_active, split = ";")[[1]]
clean_diff_active <- split_diff_active[split_diff_active != ""]
unique_diff_active <- unique(clean_diff_active)

genelist = clean_diff_active


title = paste('background:', length(unique_active), 'list:',length(unique_diff_active))


```


# Define groups of comapre enrichment
```{r}
diff_active_up <- filtered_data %>%
  filter(differential_activity == "True")%>%
  filter(logFC_derived_vs_ancestral > log2(1.25)) %>%
  pull(Gene_symbol)

combined_diff_active_up <- paste(diff_active_up, collapse = ";")
split_diff_active_up <- strsplit(combined_diff_active_up, split = ";")[[1]]
clean_diff_active_up <- split_diff_active_up[split_diff_active_up != ""]
unique_diff_active_up <- unique(clean_diff_active_up)

comp1 = "LFC above 1.25"

diff_active_down <- filtered_data %>%
  filter(differential_activity == "True")%>%
  filter(logFC_derived_vs_ancestral < -log2(1.25)) %>%
  pull(Gene_symbol)

combined_diff_active_down <- paste(diff_active_down, collapse = ";")
split_diff_active_down <- strsplit(combined_diff_active_down, split = ";")[[1]]
clean_diff_active_down <- split_diff_active_down[split_diff_active_down != ""]
unique_diff_active_down <- unique(clean_diff_active_down)

comp2 = "LFC below -1.25"

# Save the genes into the log file
write(unique_diff_active_down, file = paste0(parameters$outdir,'/',parameters$run_id,'_diff_active_down.txt'), ncolumns = 1)
write(unique_diff_active_up, file = paste0(parameters$outdir,'/',parameters$run_id,'_diff_active_up.txt'), ncolumns = 1)
write(background, file = paste0(parameters$outdir,'/',parameters$run_id,'_background.txt'), ncolumns = 1)

sink(paste0(parameters$outdir,'/',parameters$run_id,'_parameters.txt'))
print(parameters)    
sink() 

```


# Run compare test
```{r}

# compare test

compare_title = paste("up genes:",length(unique_diff_active_up), 'down genes:',length(unique_diff_active_down),' background:', length(background))
 
 GrandEnrich_compare_results = list()
 
 for(db in parameters$GrandEnrich_dbs){
   print(db)
   GrandEnrich_compare_results[[db]] = TheGrandEnrich_compare(db = db,
                       genelist1 = unique_diff_active_up,
                       genelist2 = unique_diff_active_down,
                       background = background,
                       locORser = parameters$locORser,
                       minGenes = parameters$minGenes,
                       enrORdep = parameters$enrORdep,
                       comp_main = parameters$comp,
                       comp1 = comp1,
                       comp2 = comp2,
                       uniq = parameters$uniq,
                       outdir = base_dir)
 }
   
   
   
  print('TheGrandEnrich is done!')

  HPO_compare_results = HPO_compare(
             genelist1 = unique_diff_active_up,
             genelist2 = unique_diff_active_down,
             background = background,
             minGenes = parameters$minGenes,
             TYPorALL = parameters$TYPorALL,
             HPO_dbVer = parameters$HPO_dbVer,
             systems = parameters$systems,
             bodyparts = parameters$bodyparts,
             directional = parameters$directional,
             enrORdep = parameters$enrORdep,
             comp_main = parameters$comp,
             comp1 = comp1,
             comp2 = comp2,
             locORser = parameters$locORser,
             uniq = parameters$uniq,
             outpath = base_dir
             )
  
  print('HPO enrich is done is done!')

  ORGANizer_compare_results = ORGANizer_compare(
             genelist1 = unique_diff_active_up,
             genelist2 = unique_diff_active_down,
             background = background,
             curationLevel = parameters$curationLevel,
             FreqOfPheno = parameters$FreqOfPheno,
             DBversion = parameters$DBversion,
             minGenes = parameters$minGenes,
             onlySkel = parameters$onlySkel,
             enrORdep = parameters$enrORdep,
             locORser = parameters$locORser,
             comp1 = comp1,
             comp2 = comp2,
             uniq = parameters$uniq,
             outpath = base_dir)

  
  print('ORGANizer enrich is done is done!')

  

```


# Process compare results
```{r}

HPO_compare_clean = HPO_compare_results %>%
  rename(FoldChange = Ratio)%>%
  rename(Term = Pheno)%>%
  select(-c(HPO_ID))

ORGANizer_compare_clean = ORGANizer_compare_results %>%
  rename(FoldChange = Ratio)%>%
  rename(Term = `Body Part`)%>%
  select(-c(Type))


colnames(GrandEnrich_compare_results$GO_BP)[colnames(GrandEnrich_compare_results$GO_BP) == "Ratio"] <- "FoldChange"
colnames(GrandEnrich_compare_results$KEGG)[colnames(GrandEnrich_compare_results$KEGG) == "Ratio"] <- "FoldChange"
colnames(GrandEnrich_compare_results$GAD_Disease)[colnames(GrandEnrich_compare_results$GAD_Disease) == "Ratio"] <- "FoldChange"


# Add origin column and combine
combined_compare_GrandEnrich_results <- bind_rows(list(
  GO_BP = GrandEnrich_compare_results$GO_BP,
  KEGG = GrandEnrich_compare_results$KEGG,
  GAD_Disease = GrandEnrich_compare_results$GAD_Disease,
  HPO = HPO_compare_clean,
  ORGANizer = ORGANizer_compare_clean),
  .id = "database")
  

final_compare_results = combined_compare_GrandEnrich_results%>%
  #select(database,Term,FoldChange,`P-value`,FDR)%>%
  mutate(Term = paste(database,Term),.keep = 'unused')%>%
  arrange(FDR)%>%
  arrange(desc(FoldChange))%>%
  filter(FDR<0.05)%>%
  distinct(Term,.keep_all = TRUE)

write.csv(final_compare_results,paste0(parameters$outdir,'/',parameters$run_id,'_final_results.csv'),row.names=F,col.names = FALSE)



temp = combined_compare_GrandEnrich_results%>%
  filter(database == 'HPO')%>%
  arrange(desc(FoldChange))%>%
  arrange(FDR)%>%
  select(-DB)

```


# Visualize compare
```{r}

for (database_name in unique(combined_compare_GrandEnrich_results$database)){
  final_compare_results <- combined_compare_GrandEnrich_results %>%
  filter(database == database_name)%>%
  arrange(desc(FoldChange))%>%
  arrange(FDR)%>%
  select(-DB)%>%
  filter(FDR<0.05)%>%
  mutate(text.pos = if_else(FoldChange < 1, 1, 1))
  
  if(nrow(final_compare_results)>50){
    final_compare_results = final_compare_results[1:50,]
  }
  
  
  ggplot(data = final_compare_results, aes(x =factor(Term, levels = Term[order(FoldChange, decreasing = FALSE)]),
                                 y = log10(FoldChange), fill = -log10(FDR))) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = log10(text.pos)), label = final_compare_results$Term,hjust = "left")+
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12)) +
  labs(title = paste0(compare_title,' database'),
       x = "Term",
       y = "Log Fold Change") +
  scale_fill_gradient(low = "gray", high = "red") +  # Adjusted color scale
  coord_flip()+
  theme_void()+
  theme(
    axis.line.x = element_line(),  # Add x-axis line
    axis.ticks.x = element_line(), # Add x-axis ticks
    axis.text.x = element_text()   # Add x-axis text labels
  )
  
  ggsave(paste0(base_dir, '/runs/', parameters$run_id, '/', parameters$run_id,'_',database_name, '_enrichment_results.png'), width = 20, height = 10)
}
         
                                           



ggplot(data = final_compare_results, aes(x =factor(Term, levels = Term[order(FoldChange, decreasing = FALSE)]),
                                 y = log10(FoldChange), fill = -log10(FDR))) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = log10(text.pos)), label = final_compare_results$Term,hjust = "left")+
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12)) +
  labs(title = compare_title,
       x = "Term",
       y = "Log Fold Change") +
  scale_fill_gradient(low = "gray", high = "red") +  # Adjusted color scale
  coord_flip()+
  theme_void()

ggsave(paste0(base_dir, '/runs/', parameters$run_id, '/', parameters$run_id, '_enrichment_results.png'), width = 10, height = 5)

ggplot(data = final_compare_results[1:50,], aes(x =factor(Term, levels = Term[order(FoldChange, decreasing = FALSE)]),
                                 y = log10(FoldChange), fill = -log10(FDR))) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = log10(text.pos)), label = final_compare_results$Term[1:50],hjust = "left")+
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12)) +
  labs(title = compare_title,
       x = "Term",
       y = "Log Fold Change") +
  scale_fill_gradient(low = "gray", high = "red") +  # Adjusted color scale
  coord_flip()+
  theme_void()

ggsave(paste0(base_dir, '/runs/', parameters$run_id, '/', parameters$run_id, '_enrichment_results_top50.png'), width = 10, height = 5)




```




```{r}
 # GAG_genes = GrandEnrich_results$GO_BP%>%
 #   #filter(Term =='GO:0006024~glycosaminoglycan biosynthetic process')%>%
 #   #filter(Term =='GO:0030203~glycosaminoglycan metabolic process')%>%
 #   #filter(Term =='GO:0006022~aminoglycan metabolic process')%>%
 #   filter(Term == 'GO:0006023~aminoglycan biosynthetic process')%>%
 #   pull(`Genes associated with term`)
 # 
 #   
 # 
 # GAG_genes = unlist(strsplit(GAG_genes, ","))
 # GAG_genes_pattern <- paste(GAG_genes, collapse = "|")
 #  
 # results = filtered_data %>%
 #    filter(differential_activity == "True")%>%
 #    filter(abs(logFC_derived_vs_ancestral) > log2(1.33))%>%
 #    filter(str_detect(Gene_symbol, GAG_genes_pattern))%>%
 #    as.data.frame()%>%
 #    arrange(logFC_derived_vs_ancestral)%>%
 #    select(c(oligo,logFC_derived_vs_ancestral,Gene_symbol))
 #  
 #  
 # gene_counts = results$Gene_symbol
 # gene_counts = unlist(strsplit(gene_counts, ","))
 # gene_counts <- unlist(strsplit(gene_counts, ";"))
 # 
 # gene_counts <- as.data.frame(table(gene_counts))%>%
 #   arrange(desc(Freq))
```

# Run Enrichment

```{r}
GrandEnrich_results = list()
for(db in parameters$GrandEnrich_dbs){
  GrandEnrich_results[[db]] = TheGrandEnrich_enrich(db = db,
                      genelist = genelist,
                      background = background,
                      locORser = parameters$locORser,
                      minGenes = parameters$minGenes,
                      enrORdep = parameters$enrORdep,
                      comp = parameters$comp,
                      uniq = parameters$uniq,
                      outdir = paste(parameters$outdir,'/',sep=''))
}

print('TheGrandEnrich is done!')

HPO_results = HPO_enrich(genelist = genelist,
           background = background,
           minGenes = parameters$minGenes,
           TYPorALL = parameters$TYPorALL,
           HPO_dbVer = parameters$HPO_dbVer,
           systems = parameters$systems,
           bodyparts = parameters$bodyparts,
           directional = parameters$directional,
           enrORdep = parameters$enrORdep,
           comp = parameters$comp,
           locORser = parameters$locORser,
           uniq = parameters$uniq,
           outpath = paste(parameters$outdir,'/',sep='')
           )

print('HPO enrich is done is done!')

ORGANizer_results = ORGANizer_Organize(genelist = genelist,
                   background = background,
                   outpath = paste(parameters$outdir,'/',sep=''),
                   curationLevel = parameters$curationLevel,
                   FreqOfPheno = parameters$FreqOfPheno,
                   DBversion = parameters$DBversion,
                   minGenes = parameters$minGenes,
                   onlySkel = parameters$onlySkel,
                   enrORdep = parameters$enrORdep,
                   locORser = parameters$locORser,
                   comp = parameters$comp,
                   uniq = parameters$uniq)

print('ORGANizer enrich is done is done!')

write(genelist, file = paste0(parameters$outdir,'/',parameters$run_id,'_gene_list.txt'), ncolumns = 1)
write(background, file = paste0(parameters$outdir,'/',parameters$run_id,'_background.txt'), ncolumns = 1)

write.csv(genelist,paste0(parameters$outdir,'/',parameters$run_id,'_gene_list.csv'),row.names=F,col.names = FALSE)
write.csv(background,paste0(parameters$outdir,'/',parameters$run_id,'_background.csv'),row.names=F,col.names = FALSE)

sink(paste0(parameters$outdir,'/',parameters$run_id,'_parameters.txt'))
print(parameters)    
sink()    

```

# Log

```{r}
master_log = paste(base_dir, '/enrich_master_table.csv',sep='')
log_run(master_log,
        parameters,
        data_dir,
        filterinng_description,
        background_description,
        group_description,
        gene_group_count = length(genelist),
        background_group_count = length(background))



```




# Process results


```{r}

HPO_clean = HPO_results %>%
  dplyr::rename(FoldChange = Enrichment)%>%
  dplyr::rename(Term = Pheno)%>%
  select(Term,FoldChange,`P-value`,FDR)

ORGANizer_clean = ORGANizer_results %>%
  dplyr::rename(FoldChange = Enrichment)%>%
  dplyr::rename(Term = `Body Part`)%>%
  select(Term,FoldChange,`P-value`,FDR)

# Add origin column and combine
combined_GrandEnrich_results <- bind_rows(list(
  GO_BP = GrandEnrich_results$GO_BP,
  KEGG = GrandEnrich_results$KEGG,
  GAD_Disease = GrandEnrich_results$GAD_Disease,
  HPO = HPO_clean,
  ORGANizer = ORGANizer_clean),
  .id = "database")

final_results = combined_GrandEnrich_results%>%
  select(database,Term,FoldChange,`P-value`,FDR)%>%
  mutate(Term = paste(database,Term),.keep = 'unused')%>%
  arrange(FDR)%>%
  arrange(desc(FoldChange))%>%
  filter(FDR<0.05)%>%
  distinct(Term,.keep_all = TRUE)

write.csv(final_results,paste0(parameters$outdir,'/',parameters$run_id,'_final_results.csv'),row.names=F,col.names = FALSE)



```


# Visulaize

```{r}
# Create the ggplot graph

ggplot(final_results, aes(x =factor(Term, levels = Term[order(FoldChange, decreasing = FALSE)]),
                                 y = FoldChange, fill = -log10(FDR))) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12)) +
  labs(title = title,
       x = "Term",
       y = "Log Fold Change") +
  scale_fill_gradient(low = "gray", high = "red") +  # Adjusted color scale
  coord_flip()

ggsave(paste0(base_dir, '/runs/', parameters$run_id, '/', parameters$run_id, '_enrichment_results.png'), width = 10, height = 5)



ggplot(final_results[1:30,], aes(x =factor(Term, levels = Term[order(FoldChange, decreasing = FALSE)]),
                                 y = FoldChange, fill = -log10(FDR))) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12)) +
  labs(title = title,
       x = "Term",
       y = "Log Fold Change") +
  scale_fill_gradient(low = "gray", high = "red") +  # Adjusted color scale
  coord_flip()

ggsave(paste0(base_dir, '/runs/', parameters$run_id, '/', parameters$run_id, '_enrichment_results_top30.png'), width = 10, height = 5)




```





